---
title: "Class 18 Mini Project: Investigating Pertussis Resurgence"
author: "Hubert Phan (PID: A17663330)"
format: html
---

## 1. Investigating pertussis cases by year

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r,warning=FALSE,message=FALSE}
library(datapasta)
library(ggplot2)

cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L,2022L,2023L),
  No..Reported.Pertussis.Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116,3044,7063)
)

p <- ggplot(cdc, aes(x = Year, y = `No..Reported.Pertussis.Cases`)) +
  geom_line() +
  geom_point() +
  labs(title = "Number of Reported Pertussis Cases Over Time",
       x = "Year",
       y = "Number of Reported Pertussis Cases") +
  theme_minimal()

p
```

## 2. A tale of two vaccines (wP & aP)

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r,message=FALSE,warning=FALSE}
p + 
  geom_vline(xintercept = 1946, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1996, linetype = "dashed", color = "blue") +
  annotate("text", x = 1946, y = max(cdc$No..Reported.Pertussis.Cases), label = "wP", angle = 90, vjust = -0.5, color = "red") +
  annotate("text", x = 1996, y = max(cdc$No..Reported.Pertussis.Cases), label = "aP", angle = 90, vjust = -0.5, color = "blue")
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

**Ans:** After the introduction of the aP vaccine, there was an increase in reported pertussis cases. One possible explanation for this trend is that the aP vaccine may not provide as long-lasting immunity as the wP vaccine, leading to an increase in susceptibility to pertussis over time. Additionally, changes in diagianostic practices and bacterial evolution could also contribute to the observed increase in cases.

## 3. Exploring CMI-PB data

### The CMI-PB API returns JSON data



