---
title: "Class 18 Mini Project: Investigating Pertussis Resurgence"
author: "Hubert Phan (PID: A17663330)"
format: html
---

## 1. Investigating pertussis cases by year

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r,warning=FALSE,message=FALSE}
library(datapasta)
library(ggplot2)

cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L,2022L,2023L),
  No..Reported.Pertussis.Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116,3044,7063)
)

p <- ggplot(cdc, aes(x = Year, y = `No..Reported.Pertussis.Cases`)) +
  geom_line() +
  geom_point() +
  labs(title = "Number of Reported Pertussis Cases Over Time",
       x = "Year",
       y = "Number of Reported Pertussis Cases") +
  theme_minimal()

p
```

## 2. A tale of two vaccines (wP & aP)

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r,message=FALSE,warning=FALSE}
p + 
  geom_vline(xintercept = 1946, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 1996, linetype = "dashed", color = "blue") +
  annotate("text", x = 1946, y = max(cdc$No..Reported.Pertussis.Cases), label = "wP", angle = 90, vjust = -0.5, color = "red") +
  annotate("text", x = 1996, y = max(cdc$No..Reported.Pertussis.Cases), label = "aP", angle = 90, vjust = -0.5, color = "blue")
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

**Ans:** After the introduction of the aP vaccine, there was an increase in reported pertussis cases. One possible explanation for this trend is that the aP vaccine may not provide as long-lasting immunity as the wP vaccine, leading to an increase in susceptibility to pertussis over time. Additionally, changes in diagianostic practices and bacterial evolution could also contribute to the observed increase in cases.

## 3. Exploring CMI-PB data

### The CMI-PB API returns JSON data

Use jsonlite package to read in the CMI-PB data from the API endpoint:

```{r,warning=FALSE,message=FALSE}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 

head(subject,3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r,warning=FALSE,message=FALSE}
table(subject$infancy_vac)
```

**Ans:** There are 87 aP infancy vaccinated subjects and 85 wP infancy vaccinated subjects in the dataset.

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r,warning=FALSE,message=FALSE}
table(subject$biological_sex)
```

**Ans:** There are 112 female subjects and 60 male subjects in the dataset.

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r,warning=FALSE,message=FALSE}
table(subject$biological_sex, subject$race)
```

### Side-Note: Working with dates

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r,warning=FALSE,message=FALSE}
library(lubridate)
library(dplyr)
# Use todays date to calculate age in days
subject$age <- today() - ymd(subject$year_of_birth)

ap <- subject %>% filter(infancy_vac == "aP")
wp <- subject %>% filter(infancy_vac == "wP")

avg_age_wp <- mean(time_length(wp$age, "years"), na.rm = TRUE)
avg_age_ap <- mean(time_length(ap$age, "years"), na.rm = T)

print(paste("Average wP age:", avg_age_wp))
print(paste("Average aP age:", avg_age_ap))

# Test for statistical significance
t_test_result <- t.test(time_length(wp$age, "years"), time_length(ap$age, "years"), var.equal = TRUE)
print(t_test_result)
```

> Q8. Determine the age of all individuals at time of boost?

```{r,warning=FALSE,message=FALSE}
age_at_boost <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost_years <- time_length(age_at_boost, "year")
head(age_at_boost_years)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r,warning=FALSE,message=FALSE}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

**Ans:** The histogram does show that the two groups are significantly different. Group aP is centered at a younger age than group wP, and group aP has a lower spread than group wP. Furthermore, the plot shows there is almost no overlap between the two groups.

### Joining multiple tables

Read the specimen and ab_titer tables into R and store the data as specimen and titer named data frames.

```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE) 
```

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r,warning=FALSE,message=FALSE}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```