---
title: "Class 10: Halloween Mini-Project"
author: "Hubert Phan (PID: A17663330)"
format: html
---

## Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> **Q1.** How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> **Q2.** How many fruity candy types are in the dataset?

```{r}
sum(candy[, "fruity"])
```

## What is your favorate candy?

> **Q3.** What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Haribo Gold Bears",]$winpercent
```

> **Q4.** What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

> **Q5.** What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

There is a useful skim() function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library("skimr")
skim(candy)
```

> **Q6.** Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

**Ans:** Compared to the majority of variable/columns, the n_missing column 
> **Q7.** What do you think a zero and one represent for the candy$chocolate column?
