---
title: "Class 13: Transcriptomics and the analysis of RNA-Seq data"
author: "Hubert Phan (PID: A 17663330)"
format: pdf
toc: true
---

## Bioconductor setup

Check if all required depencies are installed:

```{r, warning=FALSE, message=FALSE, results='hide'}
library(BiocManager)
library(DESeq2)
```

### DESeq2 Required Inputs

DESeq2 requires two inputs: (1) as data.frame of count data (from RNA-seq or other sequencing experiments) and (2) a second metadata data.frame which contains information about the samples. 

## Import countData and colData

Read in the counts matrix and metadata:
```{r}
# Complete the missing code
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

Take a look at the head of each:
```{r}
head(counts)
head(metadata)
```

>Q1. How many genes are in this dataset? 

```{r}
nrow(counts)
```

>Q2. How many ‘control’ cell lines do we have? 

```{r}
sum(metadata$dex == "control", na.rm = T)
```

## Toy differential gene expression

This analysis is for **demonstration** only. Note the control samples in the metadata. Find the sample id for those label control and calculate the mean counts per gene across the samples.

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

>Side-note: An alternative way to do this same thing using the dplyr package from the tidyverse. Which do you prefer and why?

**Ans:** I prefer the tidyverse method because it involves functions with names that relate to what I want to do. This makes it easier to remember which function to use.


>Q3. How would you make the above code in either approach more robust? Is there a function that could help here? 

**Ans:** The code above is not robust as there is a hard coded 4 to find `control.mean`. That would not work properly if we added more samples. A function that would help is `rowMeans` instead of `rowSums`. `rowMeans` finds the mean value of a row, making it more robust when adding or removing samples.

>Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r,warning=FALSE,message=FALSE}
library(dplyr)

# Obtain treated samples
treated <- metadata %>% filter(dex=="treated")
# Obtain treated counts
treated.counts <- counts %>% dplyr::select(all_of(treated$id))
# Calculate per row mean
treated.mean <- rowMeans(treated.counts)
head(treated.mean)
```

We will combine our meancount data for bookkeeping purposes.

```{r}
meancounts <- data.frame(control.mean, treated.mean)
colSums(meancounts)
```

>Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. 

```{r}
# base R scatterplot
plot(meancounts$control.mean, meancounts$treated.mean, xlab = "Control", ylab = "Treated")
```

> Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot? 

```{r}
library(ggplot2)

p <- ggplot(meancounts, aes(x = control.mean, y = treated.mean)) + 
  geom_point(alpha = 0.3, size = 0.8)
p
```
**Ans:** To make a scatterplot with ggplot, use the `geom_point()` function.

> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this? 

```{r,warning=FALSE}
library(scales)

p + scale_x_continuous(transform = "log2", labels = label_log(base = 2)) +
    scale_y_continuous(transform = "log2", labels = label_log(base = 2)) + 
    labs(x = "log control", y = "log Treated")
```

**Ans:** In Base R, to transform both axes include the argument `log = "xy"` within `plot()`. With ggplot, `scale_x_continuous(trans="log2")` will scale the x axis with log2 calculation.

We find candidate differentially expressed genes by looking for genes with a large change between control and dex-treated samples, focusing on their log2 values. Here we calculate log2foldchange, add it to our meancounts data.frame.

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

There are NaN and 0 values. The NaN is returned when you divide by zero and try to take the log. The -Inf is returned when you try to take the log of zero. Let’s filter our data to remove these genes. 

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

> Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

**Ans:** The purpose of `arr.ind` is to convert the typical vector output of `which()` to a matrix of coordinates (row and column indices). In this matrix, the first column contains the row indices where a zero was found. We apply unique() to these row indices because a single gene (row) might have zeros in multiple columns. Since a single zero is enough to produce invalid results (NaN and -Ind) we only need to identify that a row is "bad" once. `unique()` ensures that we have a clean set of numbers and prevents any errors from exlcuding a row multiple times.

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

```{r}
# Vector of logical operators. T means that logfc is > 2. So sum up all T

sum(up.ind)
```

> Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 

```{r}
# similar to up.ind, vector of logicals. T is values < -2 (i believe that is what question is asking)
sum(down.ind)
```
> Q10. Do you trust these results? Why or why not?

**Ans:** These results can be misleading as we have not done any statistical analysis to determine with the differences are significantly different.

## Setting up for DESeq

### Importing data

Construct a DESeqDataSet from (1) a count matrix, (2) a metadata file, and (3) a formula indicating the design of the experiment. In the metadata matrix, we are interested in the `dex` column. So, specify the design with a tilde: `design=~dex`.

Use `DESeqDataSetFromMatrix()` to build the required DESeqDataSet object.

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

## Principal Component Analysis (PCA)

Follow the DESeq recommended procedure and associated functions for PCA. First calling vst() to apply a variance stabilizing transformation and then plotPCA() to calculate our PCs and plot the results.

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex"))
```

We can also build the PCA plot from scratch using the ggplot2 package. This is done by asking the plotPCA function to return the data used for plotting rather than building the plot.

```{r}
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)

# Calculate percent variance per PC for the plot axis labels
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

## DESeq analysis

Run the DESeq pipeline on the dds object, and reassigning the whole thing back to dds, which will now be a DESeqDataSet populated with all the differential expression results.
```{r}
dds <- DESeq(dds)
```

### Getting results

```{r}
res <- results(dds)
res
```

```{r,results='hide'}
# Convert the res object to a data.frame
as.data.frame(res)
summary(res)
```

The results function contains a number of arguments to customize the results table. By default the argument alpha is set to 0.1. If the adjusted p value cutoff will be a value other than 0.1, alpha should be set to that value:

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

## Adding annotation data

Our result table so far only contains the Ensembl gene IDs. However, alternative gene names and extra annotation are usually required for informative interpretation of our results.

```{r,warning=FALSE}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

Use `columns()` on "org.Hs.eg.db" to get a list of all available key types:
```{r}
columns(org.Hs.eg.db)
```
```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")

head(res)
```

> Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res$entrez, res$uniprot and res$genename.

```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")

res$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="GENENAME",
                     keytype="ENSEMBL",
                     multiVals="first")

head(res)

```

```{r}
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
write.csv(res[ord,], "deseq_results.csv")
```

## Data Visualization

### Volcano plots

Typically these plots shows the log fold change on the X-axis, and the -log10 of the p-value on the Y-axis (the more significant the p-value, the larger the -log10 of that value will be).

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

Include guidelines to visualize cuttoffs and colors to highlght genes:
```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

To color the points we will setup a custom color vector indicating transcripts with large fold change and significant differences between conditions:

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

For even more customization you might find the EnhancedVolcano bioconductor package useful:

```{r}
library(EnhancedVolcano)

x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```