---
title: "Class 13: Transcriptomics and the analysis of RNA-Seq data"
author: "Hubert Phan (PID: A 17663330)"
format: html
---

## Bioconductor setup

Check if all required depencies are installed:

```{r,warning=FALSE}
library(BiocManager)
library(DESeq2)
```

### DESeq2 Required Inputs

DESeq2 requires two inputs: (1) as data.frame of count data (from RNA-seq or other sequencing experiments) and (2) a second metadata data.frame which contains information about the samples. 

## Import countData and colData

Read in the counts matrix and metadata:
```{r}
# Complete the missing code
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

Take a look at the head of each:
```{r}
head(counts)
head(metadata)
```

>Q1. How many genes are in this dataset? 

```{r}
nrow(counts)
```

>Q2. How many ‘control’ cell lines do we have? 

```{r}
sum(metadata$dex == "control", na.rm = T)
```

## Toy differential gene expression

This analysis is for **demonstration** only. Note the control samples in the metadata. Find the sample id for those label control and calculate the mean counts per gene across the samples.

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

>Side-note: An alternative way to do this same thing using the dplyr package from the tidyverse. Which do you prefer and why?

**Ans:** I prefer the tidyverse method because it involves functions with names that relate to what I want to do. This makes it easier to remember which function to use.


>Q3. How would you make the above code in either approach more robust? Is there a function that could help here? 



>Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)