---
title: "Class 13: Transcriptomics and the analysis of RNA-Seq data"
author: "Hubert Phan (PID: A 17663330)"
format: html
---

## Bioconductor setup

Check if all required depencies are installed:

```{r,warning=FALSE}
library(BiocManager)
library(DESeq2)
```

### DESeq2 Required Inputs

DESeq2 requires two inputs: (1) as data.frame of count data (from RNA-seq or other sequencing experiments) and (2) a second metadata data.frame which contains information about the samples. 

## Import countData and colData

Read in the counts matrix and metadata:
```{r}
# Complete the missing code
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

Take a look at the head of each:
```{r}
head(counts)
head(metadata)
```

>Q1. How many genes are in this dataset? 

```{r}
nrow(counts)
```

>Q2. How many ‘control’ cell lines do we have? 

```{r}
sum(metadata$dex == "control", na.rm = T)
```

## Toy differential gene expression

This analysis is for **demonstration** only. Note the control samples in the metadata. Find the sample id for those label control and calculate the mean counts per gene across the samples.

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

>Side-note: An alternative way to do this same thing using the dplyr package from the tidyverse. Which do you prefer and why?

**Ans:** I prefer the tidyverse method because it involves functions with names that relate to what I want to do. This makes it easier to remember which function to use.


>Q3. How would you make the above code in either approach more robust? Is there a function that could help here? 

**Ans:** The code above is not robust as there is a hard coded 4 to find `control.mean`. That would not wokr properly if we added more samples. A function that would help is `rowMeans` instead of `rowSums`. `rowMeans` finds the mean value of a row, making it more robust when adding or removing samples.

```{r}
#| eval: false

library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowMeans(control.counts)
head(control.mean)
```
>Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
library(dplyr)

# Obtain treated samples
treated <- metadata %>% filter(dex=="treated")
# Obtain treated counts
treated.counts <- counts %>% select(all_of(treated$id))
# Calculate per row mean
treated.mean <- rowMeans(treated.counts)
head(treated.mean)
```

We will combine our meancount data for bookkeeping purposes.

```{r}
meancounts <- data.frame(control.mean, treated.mean)
colSums(meancounts)
```

>Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. 

```{r}
# base R scatterplot
plot(meancounts$control.mean, meancounts$treated.mean, xlab = "Control", ylab = "Treated")
```

> Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot? 

```{r}
library(ggplot2)

p <- ggplot(meancounts, aes(x = control.mean, y = treated.mean)) + 
  geom_point(alpha = 0.3, size = 0.8)
p
```
**Ans:** To make a scatterplot with ggplot, use the `geom_point()` function.

> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this? 

```{r,warning=FALSE}
library(scales)

p + scale_x_continuous(transform = "log2", labels = label_log(base = 2)) +
    scale_y_continuous(transform = "log2", labels = label_log(base = 2)) + 
    labs(x = "log control", y = "log Treated")
```

**Ans:** In Base R, to transform both axes include the argument `log = "xy"` within `plot()`. With ggplot, `scale_x_continuous(trans="log2")` will scale the x axis with log2 calculation.

We find candidate differentially expressed genes by looking for genes with a large change between control and dex-treated samples, focusing on their log2 values. Here we calculate log2foldchange, add it to our meancounts data.frame.

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

There are NaN and 0 values. The NaN is returned when you divide by zero and try to take the log. The -Inf is returned when you try to take the log of zero. Let’s filter our data to remove these genes. 

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

> Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

**Ans:** The purpose of `arr.ind` is to convert the typical vector output of `which()` to a matrix of coordinates (row and column indices). In this matrix, the first column contains the row indices where a zero was found. We apply unique() to these row indices because a single gene (row) might have zeros in multiple columns. Since a single zero is enough to produce invalid results (NaN and -Ind) we only need to identify that a row is "bad" once. `unique()` ensures that we have a clean set of numbers and prevents any errors from exlcuding a row multiple times.

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

```{r}
# Vector of logical operators. T means that logfc is > 2. So sum up all T

sum(up.ind)
```

> Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 

```{r}
# similar to up.ind, vector of logicals. T is values < -2 (i believe that is what question is asking)
sum(down.ind)
```
> Q10. Do you trust these results? Why or why not?

