---
title: "Class 9: Halloween Mini-Project"
author: "Hubert Phan (PID: A17663330)"
format: pdf
toc: true
---

## Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> **Q1.** How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> **Q2.** How many fruity candy types are in the dataset?

```{r}
sum(candy[, "fruity"])
```

## What is your favorate candy?

> **Q3.** What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Haribo Gold Bears",]$winpercent
```

> **Q4.** What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

> **Q5.** What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

There is a useful skim() function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library("skimr")
skim(candy)
```

> **Q6.** Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

**Ans:** Compared to the majority, the winpercent looks to be on a different scale than the other variable/columns as its values are much greater than every other variable in the table.

> **Q7.** What do you think a zero and one represent for the candy\$chocolate column?

**Ans:** It is a binary indicator of whether or not the candy contains chocolate.

> **Q8.** Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy, aes(x = winpercent)) + 
  geom_histogram()
```

> **Q9.** Is the distribution of winpercent values symmetrical?

**Ans:** The distribution of winpercent values is not symmetrical. Most of the values lie within the 30-55 range despite the overall range extending from approximately 20 to 85. The distribution of data appears to be right skewed.

> **Q10.** Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
mean(candy$winpercent)

library(tidyverse) # Make sure you have loaded tidyverse for the mean function

ggplot(candy, aes(x = winpercent)) + 
  geom_histogram(binwidth = 5, color = "white", fill = "gray50") + # Added visual cleanup
  geom_vline(xintercept = 50, linetype = "dashed", color = "blue", linewidth = 1) + 
  geom_vline(xintercept = mean(candy$winpercent), color = "red", linewidth = 1.2) +
  geom_vline(xintercept = median(candy$winpercent), color = "purple", linewidth = 1.2) +
  theme_minimal()
```

**Ans:** The center of the distribution is below 50%. Since the distribution of winpercent is skewed, the median is a better representation of the center of the data, rather than mean. Because the median value of 47.8% is less than 50%, we conclude that the center of the distribution lies below 50%.

> **Q11.** On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# average the winpercent of chocolate
# average the winpercent of fruity
# compare

#create function to find average
find_average <- function(candy_type_column_name){
  #store all winpercents of candies that have candy type
   type_winpercents <- candy$winpercent[as.logical(candy[[candy_type_column_name]])]
   
   #find the average
   average_winpercent <- mean(type_winpercents, na.rm = TRUE)
   
   return(average_winpercent)
}
```

Average chocolate winpercent:

```{r}
choco_win_avg <- find_average("chocolate")
choco_win_avg
```

Average fruity winpercent:

```{r}
fruity_win_avg <- find_average("fruity")
fruity_win_avg
```

**Ans:** On average, chocolate candy is ranked higher than fruity candy.

> **Q12.** Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

**Ans:** Based on the Welch Two Sample t-test results, there is a significant difference between the win percentages of chocolate and fruity candy. We see that the p-value is much smaller than 0.05, the confidence interval indicates the true difference between the two groups is positive and substantial, and the t-test statistic is high, indicating the observed difference is far from the null hypothesis.

## Overall Candy Rankings

> **Q13.** What are the five least liked candy types in this set?

```{r}
library(tidyverse)

#arrange reg
candy %>%
  arrange((winpercent)) %>%
  head(5) %>%
  rownames()
```

> **Q14.** What are the top 5 all time favorite candy types out of this set?

```{r}
#arrange desc
candy %>%
  arrange(desc(winpercent)) %>%
  head(5) %>%
  rownames()
```

> **Q15.** Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

p <- ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()

p
```

> **Q16.** This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
p <- p + aes(winpercent, reorder(rownames(candy),winpercent))
p
```

### Time to add some useful color

Start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

Create the plot with new colors:

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> **Q17.** What is the worst ranked chocolate candy?

**Ans:** Based on the plot above, we see that the worst ranked chocolate candy is sixlets.

> **Q18.** What is the best ranked fruity candy?

**Ans:** Based on the plot above, we see the best ranked fruity candy is Starburst.

## Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> **Q19.** Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

**Ans:** Based on the plot above, we see that Reeses Minatures offer the most band for your buck.

> **Q20.** What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
top_5_expensive <- candy %>%
  arrange(desc(candy$pricepercent)) %>%
  head(5) 

top_5_expensive

least_pop <- top_5_expensive %>%
  slice_min(winpercent, n = 1) %>%
  rownames()

least_pop
```

> **Q21.** Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> **Q22.** Examining this plot what two variables are anti-correlated (i.e. have minus values)?

**Ans:** the two variables that are most anti-correlated are chocolate and fruity.

> **Q23.** Similarly, what two variables are most positively correlated?

Lets try to figure this out computationally:

```{r}
#calculate corrlation matrix
cor_matrix <- cor(candy)

# Convert the matrix to a long format to find the top pairs easily
cor_df <- as.data.frame(as.table(cor_matrix))

# Find the most positively correlated (excluding self-correlations of 1.0)
top_pos <- subset(cor_df, Var1 != Var2)
top_pos <- top_pos[order(-top_pos$Freq), ]
head(top_pos, 2)

# Find the most negatively correlated
top_neg <- cor_df[order(cor_df$Freq), ]
head(top_neg, 2)
```

**Ans:** We confirm that the most positively correlated variables chocolate and winpercent.

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

We can change the plotting character and add some color:

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

We can make a much nicer plot with the ggplot2 package but it is important to note that ggplot works best when you supply an input data.frame that includes a separate column for each of the aesthetics you would like displayed in your final plot. To accomplish this we make a new data.frame here that contains our PCA results with all the rest of our candy data. We will then use this for making plots below

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
Label the plot with ggrepel:

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

Interactive plot:

```{r}
# ggplotly() creates an HTML widget, which can’t be embedded directly into a PDF

#library(plotly)
#ggplotly(p)
```

View PC1 Loadings:

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> **Q24.** What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

**Ans:** The variables that are strongly picked up in the positive direction are fruity, hard, and pluribus. These variables logically group together to define the "non-chocolate" profile, characterizing candies that are typically fruit-flavored, hard-textured, and sold in multi-piece "pluribus" packaging. PC1 therefore separates bagged fruity candies from chocolate bar–type products based on flavor and texture.
